\markboth{BIBLIOGRAFÍA}{BIBLIOGRAFÍA}
\addcontentsline{toc}{chapter}{Bibliografía}

\begin{thebibliography}{99}
\bibitem{dagger} Dagger.io. ``Dagger Documentation | Dagger.'' Dagger.io, 2022, {\it https://docs.dagger.io}
\bibitem{blog} Dagger.io. “Dagger | Blog.” Dagger.io, 2025, {\it https://dagger.io/blog/}. Accedido el 15 de junio de 2025

\bibitem{ci} Fowler, Martin. ``Continuous Integration.'' Martinfowler.com, 18 Jan. 2024, {\it https://martinfowler.com/articles/continuousIntegration.html}.
\bibitem{img:ci} PagerDuty, Inc. ``What Is Continuous Integration?'' PagerDuty, 20 Nov. 2020, {\it https://www.pagerduty.com/resources/devops/learn/what-is-continuous-integration}.
\bibitem{img:cd} Amazon Web Services, Inc. ``¿Qué Es La Entrega Continua? – Amazon Web Services.'' Amazon Web Services, Inc., 2024, {\it https://aws.amazon.com/es/devops/continuous-delivery}.

\bibitem{cd} Fowler, Martin. ``Bliki: ContinuousDelivery.'' Martinfowler.com, 2013, {\it https://martinfowler.com/bliki/ContinuousDelivery.html}.

\bibitem{monorepo} Nx. ``Monorepo Explained.'' Monorepo.tools, 2025, {\it https://monorepo.tools}
\bibitem{kubernetes} Los autores de Kubernetes. ``Orquestación de Contenedores Para Producción.'' Kubernetes, 2025, {\it https://kubernetes.io/es}.
\bibitem{helm} Helm. ``Helm.'' Helm.sh, 2019, {\it https://helm.sh}.
\bibitem{devops} Wikipedia Contributors. ``DevOps.'' Wikipedia, Wikimedia, 1 Dec. 2019, {\it https://en.wikipedia.org/wiki/DevOps}.

\bibitem{pull} Nasser, Mohammed. ``Push vs. Pull-Based Deployments.'' DEV Community, 25 Nov. 2024, {\it https://dev.to/mohamednasser018/push-vs-pull-based-deployments-4m78}. Accedido el 14 de junio de 2025.
\bibitem{git} Git. ``Git.'' Git-Scm.com, 2024, {\it https://git-scm.com}.
\bibitem{make} Wikipedia Contributors. ``Make (Software).'' Wikipedia, 10 July 2021, en {\it https://en.wikipedia.org/wiki/Make\_(software)}.
\bibitem{just} casey. ``GitHub - Casey/Just: Just a Command Runner.'' GitHub, 2025, {\it https://github.com/casey/just}. Accedido el 14 de junio de 2025.
\bibitem{shebang} Wikipedia Contributors. ``Shebang (Unix).'' Wikipedia, 13 Aug. 2021, {\it https://en.wikipedia.org/wiki/Shebang\_(Unix)}.

\bibitem{docker} Wikipedia Contributors. ``Docker (Software).'' Wikipedia, 16 Nov. 2019, {\it https://en.wikipedia.org/wiki/Docker\_(software)}.
\bibitem{docker-compose} ``Overview of Docker Compose.'' Docker Documentation, 10 Feb. 2020, {\it https://docs.docker.com/compose}.
\bibitem{img:k8s} Kong. ``What Is Kubernetes? Examples and Use Cases.'' Kong Inc., 2024, {\it https://konghq.com/blog/learning-center/what-is-kubernetes}.
\bibitem{cluster} The Kubernetes Authors. ``Cluster Architecture.'' Kubernetes, 2025, {\it https://kubernetes.io/docs/concepts/architecture}.
\bibitem{kind} The Kubernetes Authors. ``Kind.'' Kind.sigs.k8s.io, 2025, {\it https://kind.sigs.k8s.io}.
\bibitem{kubectl} The Kubernetes Authors. ``Command Line Tool (Kubectl).'' Kubernetes, 2025, {\it https://kubernetes.io/docs/reference/kubectl}.
\bibitem{gitops} GitLab. ``¿Qué Es GitOps?'' Gitlab.com, GitLab, 9 Feb. 2022, {\it https://about.gitlab.com/es/topics/gitops}. Accedido el 20 de junio de 2025.
\bibitem{argocd} Argo Porject Authors. ``Argo CD.'' Github.io, 2025, {\it https://argoproj.github.io/cd}. Accedido el 20 de junio de 2025.
\bibitem{img:helm} Flexera. ``Kubernetes Helm: K8s Application Deployment Made Simple.'' Spot.io, 12 Sept. 2024, {\it https://spot.io/resources/kubernetes-architecture/kubernetes-helm-k8s-application-deployment-made-simple}. Accedido el 20 de junio de 2025.

\bibitem{img:dagger} Dagger. ``Introducing Dagger: A New Way to Create CI/CD Pipelines | Dagger.'' Dagger.io, 2022, {\it https://dagger.io/blog/public-launch-announcement}. Accedido el 21 de junio de 2025.
\bibitem{oci} Wikipedia Contributors. ``Open Container Initiative.'' Wikipedia, Wikimedia Foundation, 12 Nov. 2024, {\it https://en.wikipedia.org/wiki/Open\_Container\_Initiative}.
\bibitem{cue} CUE. ``The CUE Language Specification.'' CUE, 16 Apr. 2025, {\it https://cuelang.org/docs/reference/spec}. Accedido el 21 de junio de 2025.
\bibitem{sdk} Wikipedia Contributors. ``Conjunto de Herramientas de Desarrollo de Software.'' Wikipedia, 15 Aug. 2006, {\it https://es.wikipedia.org/wiki/Kit\_de\_desarrollo\_de\_software}. Accedido el 21 de junio de 2025.
\bibitem{github-actions} GitHub. ``Features $\cdot$ GitHub Actions.'' GitHub, 2025, {\it https://github.com/features/actions}.
\bibitem{go} ``The Go Programming Language.'' Go.dev, 2025, {\it https://go.dev}. Accedido el 21 junio de 2025.
\bibitem{graphql} The GraphQL Foundation. ``GraphQL: A Query Language for APIs.'' Graphql.org, 2012, {\it https://graphql.org}.

\bibitem{img:graphql} Dagger. ``Introducing the Dagger GraphQL API | Dagger.'' Dagger.io, 2022, {\it https://dagger.io/blog/graphql}. Accedido el 21 de junio de 2025.
\bibitem{daggerverse} Dagger. ``Daggerverse.'' Daggerverse.dev, 2025, {\it https://daggerverse.dev}. Accedido el 21 de junio de 2025.
\bibitem{cli} Dagger. ``ContentKeeper Content Filtering.'' Dagger.io, 2025, {\it https://docs.dagger.io/reference/cli}. Accedido el 21 de junio de 2025
\bibitem{ts} Microsoft. ``TypeScript - JavaScript That Scales.'' Typescriptlang.org, 2025, {\it https://www.typescriptlang.org}.
\bibitem{lerna} Nx. ``Lerna · a Tool for Managing JavaScript Projects with Multiple Packages.'' Lerna.js.org, 2025, {\it https://lerna.js.org}.
\bibitem{mongodb} MongoDB. ``MongoDB.'' MongoDB, 2024, {\it https://www.mongodb.com}.
\bibitem{vue} You, Evan. ``Vue.js.'' Vuejs.org, 2014, {\it https://vuejs.org}.
\bibitem{linear} LinearB, Inc. ``Cycle Time Breakdown: Tactics for Reducing PR Review Time | LinearB Blog.'' Linearb.io, 4 Aug. 2021, {\it https://linearb.io/blog/reducing-pr-review-time}. Accedido el 10 de julio de 2025.
\bibitem{age} Valsorda, Filippo. ``FiloSottile/Age.'' GitHub, 21 Apr. 2022, {\it https://github.com/FiloSottile/age}.
\bibitem{sops} getsops. ``Getsops/Sops.'' GitHub, 9 Jan. 2024, {\it https://github.com/getsops/sops}.
\bibitem{kustomize} Kubernetes SIGs. ``Kustomize - Kubernetes Native Configuration Management.'' Kustomize.io, 2025, {\it https://kustomize.io}.
\bibitem{ksops} viaduct-ai. ``GitHub - Viaduct-Ai/Kustomize-Sops: KSOPS - a Flexible Kustomize Plugin for SOPS Encrypted Resources.'' GitHub, 28 Jan. 2025, {\it https://github.com/viaduct-ai/kustomize-sops}. Accedido el 11 de julio de 2025.
\bibitem{semver} Preston-Werner, Tom. ``Versionado Semántico 2.0.0.'' Semantic Versioning, 2024, {\it https://semver.org/lang/es}.
\bibitem{bitnami} Broadcom. ``Applications - Bitnami.com.'' Bitnami.com, 2024, {\it https://bitnami.com/stacks?stack=helm}. Accedido el 12 de julio de 2025.
\bibitem{helmfile} Helmfile Authors. ``Helmfile.'' Readthedocs.io, 2025, {\it https://helmfile.readthedocs.io/en/latest}. Accedido el 12 de julio de 2025.

\bibitem{cypress} Cypress. ``JavaScript End to End Testing Framework.'' JavaScript End to End Testing Framework | Cypress.io, 2025, {\it https://www.cypress.io}.
\bibitem{dagger-custom-types} Dagger. ``Custom Types | Dagger.'' Dagger.io, 2022, {\it https://docs.dagger.io/api/custom-types}. Accedido el 13 de julio de 2025.
\bibitem{datadog} Datadog. ``Modern Monitoring \& Analytics | Datadog.'' Modern Monitoring \& Analytics, 2019, {\it https://www.datadoghq.com}.
\bibitem{prometheus} Prometheus. ``Prometheus - Monitoring System \& Time Series Database.'' Prometheus.io, 2025, {\it https://prometheus.io}.
\bibitem{horizontal} León, Manuel. ``Escalabilidad Horizontal Y Vertical: Modelos de Escalado | Arsys.'' Blog de Arsys.es, 3 July 2019, {\it https://www.arsys.es/blog/escalado-horizontal-vs-vertical}.
\bibitem{trivy} aquasecurity. ``Trivy.'' Trivy.dev, 2024, {\it https://trivy.dev/latest}.
\end{thebibliography}

